---
title: "Data_Visualization"
author: "Emily Conklin"
date: "4/30/2024"
output: html_document
---

If you don't yet have these packages installed, you may need to un-comment and run the following:

```{r setup}

#install.packages("dplyr")
#install.packages("ggplot2")
#install.packages("tidyr")
#install.packages("colorspace")
#install.packages("patchwork")
#install.packages("praise")

```

## Exercise 1: Plotting in Base R

Choose a variable of interest and plot it using only base R graphics. Use one (or more) of the following functions:
plot()
hist()
pie()
barplot()
boxplot()

See if you can figure out how to change colors, label your axes, or any other customization you'd like. 

If you're stuck, check out these resources:
https://www.theanalysisfactor.com/tutorial-part-6-basic-plotting-in-r/
https://www.datamentor.io/r-programming/plot-function/
https://www.r-graph-gallery.com/210-custom-barplot-layout.html

Or, take a peek at the key!

The starwars dataset is loaded for you, but if you'd prefer a different one,
R has lots of practice datasets - view them all with data()

```{r Base R Plots}

library(dplyr)
library(praise)

#read in starwars data and look at first few lines
data(starwars)
head(starwars)

#sample plots
hist(starwars$height)
barplot(table(starwars$gender))
plot(starwars$height, starwars$mass)

#Now you try! Note that you may run into problems using one of the
#variables in list format (films, vehicles, or starships)


#Run below
praise()

```

## Exercise 2: Basic plots in ggplot2

Below is a short exercise to start getting used to the ggplot2 syntax.

If you're stuck, take a peek at the key!

```{r ggPlot2 Plots}

library(dplyr)
library(ggplot2)
library(praise)

#read in starwars data
data(starwars)

#Below are two sample plots - a scatterplot and a barplot. 

#Try out the following for the scatterplot:
  #1) Switch the mappings for the scatterplot so that height is on the y
  #   axis and mass in on the x axis.
  #2) See if you can add a new mapping between the aesthetic size and 
  #   the variable birth_year.
  #3) Add a color mapping to the variable of your choice!
  #4) Add a shape mapping corresponding to gender

#Scatterplot
ggplot(data = starwars) + 
  geom_point(mapping = aes(x = height,
                           y = mass))

#Now, try out the following for the barplpot:
  #1) Change the x-axis mapping to a different variable of your choice.
  #2) Add a fill mapping to the variable of your choice.
  #3) Remove the legend
  #4) Add a title

#Barplot
ggplot(data = starwars) +
  geom_bar(mapping = aes(x = eye_color))

#Finally, it's time to try constructing your own ggplot!
#Using the geom_boxplot function, build a plot with the following:
  #1) Gender on the x axis
  #2) Height on the y axis
  #3) Fill corresponding to gender

#Your code here

#Run below
praise()

```

## Exercise 3: Data wrangling

Below is a short exercise on pivoting, filtering, and rearranging data before plotting.

If you're stuck, take a peek at the key!

```{r Data wrangling}

library(tidyr)
library(praise)

#read in mammal sleep dataset and look at the first few lines
data(msleep)
head(msleep)

#Try out the following:
  #1) Using filter(), remove all observations on domesticated species (conservation column)
  #2) Using pivot_longer(), convert this dataset to long format.
      #Hint: think about which columns you want to reshape ("cols" argument)
  #3) Save your filtered, pivoted dataset as a new dataset called msleep_long

#Your code here
msleep_long <- 

#Next, let's summarize!
  #1) Using group_by(), group your dataset by order
  #2) Using summarize(), find the mean sleep_total value for each order
  #3) Save your summarized dataset as a new dataset called msleep_summary

#Your code here
msleep_summary <-
  
#Let's try some merging!
  #1) Add your order-averaged values (msleep_summary) to your long-format data (msleep_long) using left_join().
  #2) Save your merged dataset as a new dataset called msleep_merged
  
#Your code here
msleep_merged <- 

#Run below
praise()
  
```

## Exercise 4: Colorspace and Patchwork

Below is a short exercise on color palettes and plot assembly.

If you're stuck, take a peek at the key!

```{r Advanced visualization}

library(colorspace)
library(patchwork)
library(praise)

#1) First, let's have some fun with colors (yay!)
#Run the line below and choose a palette from the popup.

pal <- choose_palette()
pal_vals <- pal(8) #get 8 colors from that palette

#And test it out on a demo plot - try different plot types!
demoplot(pal_vals, "bar")

#Next, we're going to check to see if our palette is color-palette friendly. 
#To get you started:

deutan_sim <- deutan(pal_vals)
demoplot(deutan_sim, "bar")

#Repeat with protan() and tritan()
#If you can't distinguish all colors from each other, pick a different palette.

#Your code here

#2) Next, build two ggplots of your choice using the 'msleep' dataset (or your merged version)
# If you're not sure where to start, try a scatterplot (geom_point) or barplot (geom_col or geom_bar)
# Save these plots as two objects called 'p1' and 'p2'
# You can always add a second color palette!

#Your code here
p1 <- 

p2 <- 
  
#3) Finally, let's arrange your plots using patchwork! Try out:
  #1) Arranging the plots in one row
  #2) Arranging the plots in one column
  #3) Add tag annotations to your plot
  #4) Apply theme_bw() to both plots
  
#Your code here
  
#Run below
praise()  
  
```


