---
title: "Debugging_Exercises"
author: "Emily Conklin"
date: "5/1/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("reprex")
#install.packages("ggplot2")
#install.packages("dplyr")

```

## Exercise 1: Data Types and Structures

Some empty data structures are provided - fill them with data and follow the commented instructions

```{r}

#First, create a vector named int_vector that holds 5 numbers
#Next, create a vector nmed char_vector that holds 5 letters
#Remember to put your letters in quotes ""

int_vector <- c(1, 2, 3, 4, 5)
char_vector <- c("a", "b", "c", "d", "e")

#Examine each vector using typeof(). 
typeof(int_vector)
typeof(char_vector)

```

```{r}

#Next, we'll build a dataframe containing int_vector and char_vector from the last step
#This dataframe has two variables: "int" and "char"

df <- data.frame("int" = int_vector, "char" = char_vector)

#Examine your dataframe with class() and typeof()
class(df)
typeof(df)

#You can select a specific column using $: dataframe_name$column_name
#Use class() and typeof() on each column in your dataframe
class(df$int)
class(df$char)
typeof(df$int)
typeof(df$char)

```

```{r}

#Next, try to create a new vector (comb_vector) containing 3 numbers and 3 letters
#What happens? 

comb_vector <- c(1, 2, 3, "a", "b", "c")
comb_vector #prints out contents of comb_vector

#What does this vector look like with typeof()?
typeof(comb_vector)

```

```{r}
#Consider the following lines of code before you run them - what do you expect?

typeof(TRUE)
typeof("TRUE")
typeof(NA)
typeof(NaN)
typeof("NA")

```

```{r}

#Finally, build a list containing 'comb_vector' and 'df' from earlier in the exercise

comb_list <- list(comb_vector, df)

#What are the outputs of class() and typeof() for comb_list?
class(comb_list)
typeof(comb_list)

#You can select a specific item in your list using [[]]: list_name[[1]]
#What are the outputs of class() and typeof() for different items in your list?
class(comb_list[[1]])
typeof(comb_list[[1]])
class(comb_list[[2]])
typeof(comb_list[[2]])

```

## Exercise 2: Stack Trace

Stack traces are a useful way to keep track of variables, especially when your code
isn't behaving as predicted, or when you're trying to make sense of someone else's 
code. To do a stack trace by hand, we can set up our variables on the left and the
"stack" (memory) on the right. 

We'll read it bottom to top, and "push" new variables to the top of the stack as we go.

```{r Stack Trace}

# We'll do this stack trace as a group, reading through the code and adding 
# to our stack trace as we go.

# !! DON'T RUN THIS CODE UNTIL AFTER THE STACK TRACE !!

fruits <- c("papaya", "breadfruit")
nums <- c(2, 4, 7)

for (i in 1:2){
  
  j <- i + nums[i]
  
  statement <- paste("We have", i, fruits[1], "and", j, fruits[2])
  print(statement)
  
}

```

## Exercise 3: Reprex

Here, we have a chunk of code that produces an error. Your job is to build a minimum reproducible example, formatted with the reprex library. Once you have your reprex, see if you can figure out what's causing the error.

Remember, your goal when building your reprex is to find the minimum amount of code that still reproduces the error - get rid of any packages, data, or function calls you don't need.

The library "reprex" is handy - once you build a reprex, you can copy it to your clipboard and run the command reprex(). This will do two things:

1) Test your code to see if it's actually reproducible (and produce other errors if not)
2) Format it nicely, so it can be easily pasted to StackOverflow or somewhere similar

```{r Original Code}

##### ORIGINAL CODE HERE

#load libraries
library(dplyr)
library(reprex)
library(ggplot2)

#load data
data(mtcars)
data(starwars)

#sort mtcars by horsepower
mtcars <- mtcars %>%
  arrange(hp)

#find average # cylinders for cars with at least 100 horsepower
avg_cylinders <- mtcars %>%
  mutate(car_owner = starwars$name[1:32]) %>%
  mutate(make = row.names(mtcars)) %>% 
  filter(hp >= 100) %>%
  summarize(mean(cylinders))
```  

```{r Reprex}

##### YOUR REPREX HERE
library(dplyr)

#load data
data(mtcars)

#find average # cylinders for cars with at least 100 horsepower
avg_cylinders <- mtcars %>%
  summarize(mean(cylinders)) #ERROR HAPPENS HERE

# Copy your reprex to your clipboard, then run the next lines to test + format
library(reprex)
reprex()

```

## Exercise 4: Debugging

We're going to try out the simplest R debugging tool, traceback(). This example comes from the Advanced R site if you're interested in a more detailed debugging tutorial:
https://adv-r.hadley.nz/debugging.html

```{r}

#Here’s a simple example: you can see that f() calls g() calls h() calls i(), which checks if its argument is numeric:Function f calls function g 

f <- function(a) g(a)
g <- function(b) h(b)
h <- function(c) i(c)
i <- function(d) {
  if (!is.numeric(d)) {
    stop("`d` must be numeric", call. = FALSE)
  }
  d + 10
}

#When we run this function, we get an error:
f("a")

#Call "traceback" to see the path to the error:
traceback()

```

